---
title: "Darwin's data"
output: markdowntemplates::minimal
---

```{r setup, include=FALSE, message=F, warning=F, error=F, comment=NA}
knitr::opts_chunk$set(echo=T, eval=T, message=F, warning=F, error=F, comment=NA)
```

A scientific study consists of document(s) specifying a population, question, hypothesis, 
experimental design, experimenter, data, analysis plan, analyst, code, parameter estimates, and claims about the parameter estimates (Patil et al., 2016)

Charles Darwin collected data over a period of years on the heights of Zea mays plants. 

### Population

Zea mays plants

### Darwin's hypothesis

Height of a plant depends on the type of fertilization. 

### Experimental Design

The plants were descended from the same parents and planted at the same time. Half of the plants were *self-fertilized*, and half were *cross-fertilized*, and the purpose of the experiment was to compare their *heights* (measured in eighths of an inch). To this end Darwin planted them in pairs in different pots.

The focus of interest is the relation between the height of a plant and something that can be controlled by the experimenter, namely whether it is self or cross-fertilized. This means that you can regard the height as random with a distribution that depends on the type of fertilization, which is fixed for each plant. 

Note that in order to minimize differences in humidity, growing conditions, lighting, etc. Darwin had decided to plant the seeds in pairs in the same pots.
The height of a plant would therefore also depend on these factors, which are not of interest, not only on the type of fertilization. 

### Data

```{r, echo=FALSE}
library(SMPracticals)
data(darwin)
darwin_pair = data.frame(Pot = darwin[darwin$type=="Cross",1],
           Cross = darwin[darwin$type=="Cross",4],
           Self = darwin[darwin$type=="Self",4])
darwin_pair
```


## Questions


1. Does height depend on the type of fertilization? Compute the $p$-value for testing the null hypothesis of no fertilization effect. 

2. Could you estimate the height difference, and assess the uncertainty of the estimate? Compute the confidence intervals for the fertilization effect. 


## Task

Answer 1. and 2. by specifying

*  **Analysis Plan**: Define a statistical model, and be very specific about the required assumptions. You can use different models to answer 1 and 2. 

* **Code**: Write the code to get the results. Your code must be replicable. 

* **Claim:** Briefly comment your results in 1. and 2.


## References

Patil, Peng, and Leek (2016) A statistical definition for reproducibility and replicability. BioRxiv.
