---
title: "Darwin's data"
output: markdowntemplates::minimal
---

```{r setup, include=FALSE, message=F, warning=F, error=F, comment=NA}
knitr::opts_chunk$set(echo=T, eval=T, message=F, warning=F, error=F, comment=NA)
```


```{r, echo=FALSE}
library(SMPracticals)
data(darwin)
darwin_pair = data.frame(Pot = darwin[darwin$type=="Cross",1],
           Cross = darwin[darwin$type=="Cross",4],
           Self = darwin[darwin$type=="Self",4])
```

## Reference for the solution(s)

A.C. Davison (2003) Statistical Models. Cambridge University Press. Examples 1.1, 3.11, 3.16, 7.24, 7.28, 8.13, 8.19


## Galton's model

Galton assumed a model where the height of a self-fertilized plant is
$$
Y = \mu + \sigma \varepsilon
$$
and of a cross-fertilized plant is
$$
X = \mu + \theta + \sigma \epsilon
$$
where $\mu$, $\theta$ and $\sigma$ are unknown parameters, and $\varepsilon$ and $\epsilon$ are independent random variables with mean zero and unit variance. 

Observations from self-fertilized plants $Y_1,\ldots,Y_{15}$ are i.i.d. as $Y$, and observations from cross-fertilized plants $X_1,\ldots,X_{15}$ are i.i.d. as $X$. 

The comparison between two independent samples are usually visualized by a boxplot:

```{r}
library(ggpubr)
ggboxplot(darwin, x = "type", y = "height",color = "type")
```

If we assume that $\epsilon_i$ and $\varepsilon_i$ have a $N(0,1)$ distribution, we can use a two-sample t test

```{r}
t.test(height  ~ type, data=darwin, var.equal=TRUE)
```

From the two-sample t test we obtain a point estimate of $\hat{\theta}=20.19-17.57= 2.62$ along with a 95\% confidence interval
$(0.417, 4.816)$ not including 0. The answer to Q1 is there is a significant height increase - estimated on average 2.62 eighths of an inch for cross-fertilized plants - and we reject $H_0:\theta=0$ with $p$-value $= 0.021$ (two-sided alternative). The answer to Q2 is that the average height increase is between $(0.417, 4.816)$ with 95\% confidence.

Alternatives: Welch's test

```{r}
t.test(height  ~ type, data=darwin)
```

or by including pot's effect

```{r}
summary(lm(height  ~ 0 + pot + type, data=darwin))
```


## Fisher's model

In order to minimize differences in humidity, growing conditions, and lighting, Darwin had taken the trouble to plant the seeds in pairs in the same pots.

Comparison of different pairs would therefore involve
these differences, which are not of interest, whereas
\emph{comparison within pairs} would depend only on the type of fertilization. 

Fisher considered the model
$$Y_i = \mu_i + \sigma \varepsilon_{i}, \quad X_i = \mu_i + \theta + \sigma \epsilon_{i}, \quad i = 1,\ldots,n$$

The parameter $\mu_i$ represents the effects of the planting conditions for the $i$th pair,
and $\varepsilon_i$ and $\epsilon_i$ are independent random variables with mean zero and unit variance. 

The $\mu_i$ could be eliminated by using the differences
$$D_i = X_i - Y_i$$
which have mean $\theta$ and variance $2\sigma^2$. 

The comparison within pairs can be visualized by the following plot:

```{r fig.cap="Darwin data: difference in height for each pair"}
ggline(darwin, x = "type", y = "height", group="pair")
```

If we assume that $\epsilon_i$ and $\varepsilon_i$ have a $N(0,1)$ distribution, we can use a \emph{paired t test}, or one-sample t test for the difference 

```{r}
differences <- apply(darwin_pair[,3:2],1,diff)
t.test(differences)
```

From the one-sample t test we obtain a point estimate of $\hat{\theta}= 2.62$ (exactly as before) along with a 95\% confidence interval
$(0.004, 5.24)$. The answer to Q1 is that although we obtain the same estimated height increase of 2.62 for cross-fertilized plants, there more uncertainty than before in concluding that it is significantly different from zero ($p$-value = 0.0497, two-sided alternative). The answer to Q2 is that the average height increase is between $(0.004, 5.24)$ with 95\% confidence. A long interval close to zero is not surprising given the small sample size ($n=15$).





As an alternative: consider the analysis of variance with, in order: no differences among pairs and no difference
between cross-fertilization and self-fertilization; no differences among pairs but an
effect of fertilization type; differences among the pots and an effect of fertilization
type; and differences among the pots and among the pairs and an effect of fertilization
type:

```{r}
summary(aov(lm(height ~ type + pot +  pair, darwin)))
```



## Distribution-free model

Consider a more general matched pair model for the Darwin data:
$$Y_i = \mu_i + \sigma_i\varepsilon_{i}, \quad X_i = \mu_i + \theta + \tau_i\epsilon_{i}, \quad i = 1,\ldots,n$$
with  $\varepsilon_i$ and $\epsilon_i$ are independent random variables with continuous distribution and mean zero and unit variance.


The height differences may be written as
$$D_i = \theta + (\tau_i\epsilon_{i} - \sigma_i\varepsilon_{i})$$
Assume that $\varepsilon_{i}$ and $\epsilon_{i}$ are independent and symmetrically distributed around 0 ( e.g. Uniform$(-i,i)$ ),  then $D_i$ is symmetrically distributed around $\theta$, i.e. $$D_i - \theta \stackrel{d}{=} \theta - D_i \qquad i=1,\ldots,n$$

If we assume that $\epsilon_i$ and $\varepsilon_i$ are independent and symmetrically distributed around 0, then $D_i$ is symmetrically distributed around $\theta$ and we can test $H_0:\theta=0$ by a nonparametric or distribution-free test, the \emph{sign test} for the difference (number of successes = 13, number of trials = 15, p-value = 0.007385 [two-sided], where success means that in a given pair the height of cross-fertilized > self-fertilized, and trials are the number of pairs). 

```{r}
binom.test(x=13, n=15, p=0.5, alternative="two.sided")
```

The sign test provides significant evidence that $\theta\neq 0$ ($p$-value = 0.007385, two-sided alternative) with little assumptions. 
The probability for cross-fertilized plant of being higher than its self-fertilized counterpart is estimated $86\%$ with a 95\% confidence interval of $(0.59, 0.98)$.




