---
title: "Solution HW4"
output: markdowntemplates::minimal
---

```{r setup, include=FALSE, message=F, warning=F, error=F, comment=NA}
knitr::opts_chunk$set(echo=T, eval=T, message=F, warning=F, error=F, comment=NA)
```

Let $M=\{1,\ldots,m\}$ be the indexes of the null hypotheses $H_i$ with $p$-values $p_i$, $i\in M$.
Assume $p_1,\ldots,p_m \stackrel{i.i.d}{\sim}U(0,1)$. 
For any $\alpha \in (0,1]$, let $S=\{i\in M: p_i \leq \alpha \}$ (with $S=\emptyset$ if $p_i> \alpha$ for all $i \in M$). Note that $|S| \sim \mathrm{Binomial}(m,\alpha)$.

### Bonferroni

Let $R = \{i \in S: p_i \leq \alpha/|S|\}$ with $R=\emptyset$ if $S =\emptyset$. Note that $p_i|i\in S \sim U(0,\alpha)$. 
Then we have
\begin{align*}
\mathrm{FWER} = & \sum_{k=1}^{m}\mathrm{P}( \cup_{i\in S} \{p_i \leq \alpha/k\}| |S|=k)\mathrm{P}(|S|=k)\\
= & \sum_{k=1}^{m} (1-(1-1/k)^k){ m \choose k }\alpha^k (1-\alpha)^{(m-k)}
\end{align*}



### Benjamini-Hochberg


Let $R$ be the indexes of the rejections of the Benjamini-Hochberg procedure at level $\alpha$ applied to the selected hypotheses  $H_i, i \in S$.
Note that $R=S$ and $|S| \sim \mathrm{Binomial}(m,\alpha)$. 
Then we have
\begin{align*}
\mathrm{FDR} = & \mathrm{P}(S \neq \emptyset)\\
= & 1 - (1-\alpha)^m
\end{align*}

```{r, echo=F}
ms = 1:1000
alpha = 0.05
FWER = sapply(ms, function(m)
    sum(
      sapply(1:m, function(k)
      (1-(1-1/k)^k)*dbinom(k,size=m,prob=alpha)
      )
    )
  )
FDR = 1 - (1-alpha)^ms
plot(ms,FDR, type="l", ylim=c(0,1), xlab="m", ylab="Error rate", col=4)
lines(ms,FWER, col=2)
legend(800, 0.2, legend=c("FWER", "FDR"),
       col=c("red", "blue"), cex=0.8, lty=1)
```
