---
title: "Cholesterol data"
output: markdowntemplates::minimal
---

```{r setup, include=FALSE, message=F, warning=F, error=F, comment=NA}
knitr::opts_chunk$set(echo=T, eval=T, message=F, warning=F, error=F, comment=NA)
```


This example is a hypothetical one. Twelve men are recruited from among those attending a fitness clinic and asked to participate in an experiment to establish whether eating fish (but not meat) results in lower plasma cholesterol concentrations than eating meat (but not fish). 

The subjects are randomly allocated to the fish-eating and meat-eating regimens. At the end of one year their plasma cholesterol concentrations are measured. These are:

```{r, echo=F}
rm(list=ls())

X = c(5.42, 5.86, 6.16, 6.55, 6.80, 7.00, 7.11)
Y = c(6.51, 7.56, 7.61, 7.84, 11.50)
list(fish=X,meat=Y)
Z = c(X,Y)
n.x = length(X)
n.y = length(Y)
n.z = n.x + n.y
```


Permutation null distribution of the two-sample t statistic

```{r, echo=F}
library(combinat)
perms = combn(1:n.z,n.x)
t.stat = apply(perms, 2, 
              function(perm) t.test(Z[-perm],Z[perm],var.equal=T,                            alternative="greater")$statistic
              )
plot(table(round(t.stat,3)), 
     xlab = "t statistic",
     ylab = "frequency")
points(t.stat[1],0,pch=4,col="red")
```

Permutation $p$-value

```{r, echo=F}
mean( t.stat >= t.stat[1] ) 
```


Permutation null distribution of the Wilcoxon-Mann-Whitney statistic

```{r, echo=FALSE}
R = rank(Z)
w.stat = apply(perms, 2, 
              function(perm) sum(R[-perm])
              )
plot(table(w.stat), 
     xlab = "sum of ranks",
     ylab = "frequency")
points(w.stat[1],0,pch=4,col="red")
```

Permutation $p$-value

```{r, echo=F}
mean( w.stat >= w.stat[1] )
wilcox.test(Z[(n.x+1):n.z],Z[1:n.x], 
            alternative="greater",
            exact=TRUE)
```